---
title: "Data 612 - Project 2"
author: "Habib Khan, Vijaya Cherukuri"
date: "June 15, 2020"
output: 
  html_document:
    toc: yes
    toc_float: yes
---

# Introduction

Start with an existing dataset of user-item ratings, such as our toy books dataset, MovieLens, Jester or another dataset of your choosing. Implement at least two two of these recommendation algorithms:

1- Content-Based Filtering <br>
1.1 User-User Collaborative Filtering <br>
1.2 Item-Item Collaborative Filtering



```{r, message=FALSE, warning=FALSE}
# Loading libraries
library(knitr)
library(kableExtra)
library(tidyverse)
library(recommenderlab)
```

# Overview of data

Although there are lots of dataset for recommender systems but for sake of simplicity I will include MovieLens that is built-in data. 

```{r, message=FALSE, warning=FALSE}
# Loading the MovieLense dataset
set.seed(233)
data("MovieLense")

# Creating a seperate dataframe
movies <- MovieLense@data

# Data structure
dim(movies)
```

The MovieLense dataset contains reviews from 943 users and 1664 movies.

# Data Exploration

Let's explore the distribution of user's review via barchart and heatmap

```{r}
# Creating barcharts of ratings

movies %>% as.vector() %>%  as.tibble() %>% filter_all(any_vars(. != 0)) %>% 
  ggplot()+ geom_bar(aes(value))+ labs(title="Movie Ratings by users", x="Ratings", y="Numbers")+ theme_classic()
```

Most of the reviews consist around 4 and somewhere 3. Now let's create heatmap to see overall picture of the dataset.

```{r}
# Creating heatmap for the entire dataset
image(movies, main="Heatmap - Ratings")
```

```{r}
# Mostly viewed movies by users
table <- data.frame(
  movie= names(colCounts(MovieLense)),
  views= colCounts(MovieLense)
)

# Mostly viewed movies
table %>% head() %>% kable() %>% kable_styling()

# Creating graph for mostly viewed movies
ggplot(table[1:10, ], aes(reorder(x=movie, views), y=views))+geom_bar(stat="identity")+theme(axis.text.x = element_text(angle=45, hjust=1))+labs(title="Mostly viewed movies", x="Movies", y="Views")
```


# Data Preparation

Data preparation is a significant step to create an accurate prediction and recommendations. According to Gorakala and Usuelli(2015), movies that have been viewed only a few times and users who rated only a few movies that's why it creates biasness in the result. To avoid biasness, I will select users who rated at least 50 movies and watched at least 100 times. Another requirement is also needed that is NORMALIZATION which I won't do as Recommender function does it by itself.

```{r}
# Selecting data 
movies_s <- MovieLense[rowCounts(MovieLense) > 50, colCounts(MovieLense) > 100]

# Creating train and test dataset
which_movies <- sample(x=c(TRUE, FALSE), size= nrow(movies_s), replace = TRUE, prob=c(0.8, 0.2))
movies_train <- movies_s[which_movies, ] # Training
movies_test <- movies_s[!which_movies, ] # Testing

```

# Item based Collaborative Filtering

Item based collaborative filtering is type of recommender systems in which similarities between the items are calculated for recommendation. Let's create similary matrix to see the similarity between the items. I will create similarity index for 10 movies

```{r}
# Create similarity index for 10 movies
sim_item <- similarity(MovieLense[, 1:4], method="cosine", which="items")
image(as.matrix(sim_item), main="Item similarity")

# Creating recommender system - item based collaborative filtering
item_model <- Recommender(movies_train, method="IBCF", parameter= list(k=30))
model_detail <- getModel(item_model)

# Creating heatmap for few rows and columns
image(model_detail$sim[1:20, 1:20], main="Heatmap for few rows and columns")

# Creating predictions
train_pred <- predict(object= item_model, newdata = movies_test, n=10)
train_matrix <- sapply(train_pred@items, function(x) {colnames(movies_s)[x]})
train_matrix[, 1:5] %>% kable() %>% kable_styling()
```


# User based collaborative filtering

User based collaborative filtering is a recommender system that measures similarity between users. Below, I will create a similarity matrix and then create a model to predict movies based on similarity among the users.

```{r}
# Creating similarity index and image
sim_users <- similarity(MovieLense[1:4, ], method="pearson", which="users")
image(as.matrix(sim_users), main="Similarity among users")

# Creating a recommender model - user based collaborative filtering
sim_model2 <- Recommender(movies_train, method="UBCF", parameter=list(k=25))
model_detail2 <- getModel(sim_model2)

# Calculate predictions
train_ub_pred <- predict(object=sim_model2, newdata= movies_test, n=10)
train_ub_matrix <- sapply(train_ub_pred@items, function(x) {colnames(movies_s)[x]})
train_ub_matrix[, 1:5] %>% kable() %>% kable_styling()

```





# Reference

https://www.packtpub.com/big-data-and-business-intelligence/building-recommendation-system-r
